The zoon R package for reproducible and shareable species distribution modelling

### 2016-12-23 ms.Rmd
plotting for Example 3, and tidying up first section

@@

### Example workflow

The following workflow demonstrates use of the chains and lists to fit three different SDMs to a presence-background dataset for a UK mosquito species, after standardising the covariates.
The results of the three models are then returned in Shiny app allowing the user to interactively explore the occurrence and covariate data, model summaries and prediction maps.

```{r plotworkflow, eval = TRUE, echo = TRUE, fig.height = 4, fig.width = 9, dpi = 300}
work <- workflow(occurrence = UKAnophelesPlumbeus,
                 covariate  = UKBioclim,
                 process    = Chain(Background(n = 500),
                                    StandardiseCov),
                 model      = list(MaxEnt,
                                   GBM,
                                   RandomForest),
                 output     = Appify)
```

@@

We demonstrate the zoon R package by recreating two SDM analyses from published research articles as zoon workflows and showing how these workflow objects can easily be extended with new modules.
Workflow objects created by these analyses can be accessed at [`figshare.com/articles/zoon_applications_paper_workflows`](http://figshare.com/articles/zoon_applications_paper_workflows).
We encourage readers to download, interrogate and alter these workflows for themselves.
Full code and metadata for all of the modules used in the examples below can be found at [`github.com/zoonproject/modules/R`](https://github.com/zoonproject/modules/R)

@@

Next, we update the workflow to produce an interactive map enabling anyone to inspect the data and predictions on a zoomable map, and to inspect the response curves of the fitted model. These outputs are shown in Figure This map is shown in Figure 2, panels B and C.

@@

```{r feng_papes_plots, cache = FALSE, echo = FALSE, out.width = "450px", fig.cap = "Outputs of the workflow objects `FengPapes' and `FengPapesUpdate'. A) Map of the presence data (red), background data (black) and the MaxEnt relative probability of occurrence generated by the `PrintMap' module in the workflow `FengPapes'. White areas are masked due to being in the sea or outside the MESS mask. B) A response curve produced by the `ResponseCurve' module for the first covariate, bio1 in the workflow `FengPapesUpdate'. C) A screenshot of the interactive map produced by the `InteractiveMap' modules in the workflow `FengPapesUpdate', displaying raw occurrence data and predicted distribution over a global map, allowing users to interactively explore their results."}
knitr::include_graphics("figs/combinefengupdate-1.png")
```

@@

This module can now be shared so that others can use it in their zoon workflows. The best way of sharing it is via the zoon modules repository. Modules can uploaded there, via the module submission system at [URL TBD].

#### Example 3. Evaluating MaxLike

@Royle2012 proposed a model for fitting SDMs to presence-only data, which they suggest is able to estimate the absolute (rather than relative) probability of species presence in each grid cell.
This model and the associated claims have been the subject of several research articles evaluating the model against both real-world and simulated datasets [@Fitzpatrick2013;@Merow2014;@Hastie2013;@Phillips2013].

In this example we illustrate how zoon can help contribute to discussions such as these by enabling rapid evaluation of a new method against a variety of datasets and test procedures.

For this example, it was necessary to create three new modules: two to load the presence/absence and presence-only Carolina Wren datasets from @Royle2012; and one to implement the MaxLike model.

<!-- Possible experiments:
* evaluating with different criteria (e.g. deviance or pseudo-r-squared which measure calibration capacity)
* fitting the same models with GBIF data and evaluating against the BBS PA data,
* fitting/evaluating on spatially-stratified holdout data
* fitting the models with other species
  (choose only one of these)
  -->

All of these have now been uploaded to the zoon modules repository under the names `CarolinaWrenPO`, `CarolinaWrenPA`, `CarolinaWrenRasters` and `MaxLike`.

What the workflow looks like:

```{r merowworkflowtemp, dpi = 600, eval = TRUE}
Merow_Silander <- workflow(
    occurrence = CarolinaWrenPO,
    covariate = CarolinaWrenRasters,
    process = Background(n = 10000),
    model = list(LogisticRegression, MaxEnt, MaxLike),
    output = Chain(PerformanceMeasures, PrintMap(points = FALSE)))
```

```{r merowworkflow,dpi = 600, eval = FALSE}

Merow_Silander <- workflow(
    occurrence = CarolinaWrenPO,
    covariate = CarolinaWrenRasters,
    process = Chain(Transform(trans = 'square',
                                 replace = FALSE), 
                       Background(n = 10000)),
    model = list(LogisticRegression, MaxEnt, MaxLike),
    output = Chain(PerformanceMeasures, PointBiserialCorr, PrintMap(points = FALSE), Coefficients))
```

So we ran it again with disc-based spatial stratification:

```{r merowChangeWorkflow, dpi = 600, eval = FALSE}
Merow_Silander_Spatial <- ChangeWorkflow(
    merow_and_silander,
    process = Chain(..., PartitionDisc))
```

>>>>

The following code uses `workflow` to run simple a presence-background SDM for a mosquito species in the UK, fitting a MaxEnt model with default settings and 500 randomly placed background points.

```{r plotworkflow, eval = TRUE, echo = TRUE, fig.height = 4, fig.width = 9, dpi = 300}
mosquito1 <- workflow(occurrence = UKAnophelesPlumbeus,
                      covariate  = UKBioclim,
                      process    = Background(n = 500),
                      model      = MaxEnt,
                      output     = InteractiveMap)
```

@@

###

The following workflow uses a chain to standardise the covariate rasters as well as generating background records, and uses a list to fit three different models: MaxEnt, Boosted Regression Trees (GBM) and RandomForest.
The results of the three models are then returned in three Shiny apps allowing the user to interactively explore the occurrence and covariate data, model summaries and prediction maps.

```{r workflow2, eval = TRUE, echo = TRUE, fig.height = 4, fig.width = 9, dpi = 300}
mosquito2 <- workflow(occurrence = UKAnophelesPlumbeus,
                      covariate  = UKBioclim,
                      process    = Chain(Background(n = 500),
                                         StandardiseCov),
                      model      = list(MaxEnt,
                                        GBM,
                                        RandomForest),
                      output     = Appify)
```

@@

Next we demonstrate how zoon can be used in practice for running SDMs, creating and sharing new modules, and performing a methods comparison. The workflow objects created by these analyses can be accessed at [`figshare.com/articles/zoon_applications_paper_workflows`](http://figshare.com/articles/zoon_applications_paper_workflows).
We encourage readers to download, interrogate and alter these workflows for themselves.

@@

Next, we update the workflow to produce an interactive map enabling anyone to inspect the data and predictions on a zoomable map, and to inspect the response curves of the fitted model. These outputs are shown in Figure 2, panels B and C.

@@

```{r feng_papes_plots, cache = FALSE, echo = FALSE, out.width = "450px", fig.cap = "Outputs of the workflow objects `FengPapes' and `FengPapesUpdate'. A) Map of the presence data (red), background data (black) and the MaxEnt relative probability of occurrence generated by the `PrintMap' module in the workflow `FengPapes' which encodes a whole published analysis. White areas are masked due to being in the sea or outside the MESS mask. B) A response curve produced by the `ResponseCurve' module for the first covariate, bio1 in the workflow `FengPapesUpdate', which modifies the original analysis workflow. C) A screenshot of the interactive map produced by the `InteractiveMap' modules in the workflow `FengPapesUpdate', displaying raw occurrence data and predicted distribution over a global map, allowing users to interactively explore their results. Any SDM analysis distributed as a zoon workflow can be easily be explored and scrutinized by modifying its output modules using the function `ChangeWorkflow'."}
knitr::include_graphics("figs/combinefengupdate-1.png")
```

@@

This module can now be shared so that others can use it in their zoon workflows.
Modules can be uploaded to the zoon modules repository via the online submission system at [URL TBD].

#### Example 3. What makes MaxEnt effective

The popular MaxEnt SDM model has recently been shown to be equivalent to a Poisson point process model [@Renner2013], and to be closely approximated by a logistic regression model with weights applied to background datapoints [@Warton2010; @Fithian2013; @Renner2015].
Given this close correspondence between MaxEnt and logistic regression, @Renner2013 and others have suggested that MaxEnt's superior predictive performance [@Elith2006] is most likely due to the array of features (candidate covariate transformations) it constructs and its use of regularisation to prevent overfitting.

zoon enables us to investigate this hypothesis by fitting MaxEnt models with and without regularisation and feature construction.
We fit MaxEnt with both the usual java implementation as well as the downweighted logistic regression approximation provided in the `maxnet` R package [@maxnet].
We also compare these models with a standard logistic regression model which does not apply the downweighting step.
The following workflow fits these models to a previously published presence-background data on the Caroline wren in the USA [@Royle2012; @maxlike], maps their predictions (Figure 3) and evaluates the predictive performance of each model against the full presence-absence dataset by AUC.

```{r maxnet_comparison, dpi = 600, fig.show = "hide", eval = TRUE, fig.height = 7, fig.width = 9}
maxnet_comparison <- workflow(
  occurrence = CarolinaWrenPO,
  covariate  = CarolinaWrenRasters,
  process    = Chain(SubsampleOccurrence(500),
                     Background(n = 10000),
                     CarolinaWrenValidation),
  model      = list(MaxEnt(args = 'threshold=false'),
                    MaxNet,
                    MaxNet(regmult = 0.005),
                    MaxNet(features = 'l'),
                    MaxNet(regmult = 0.005, features = 'l'),
                    LogisticRegression),
  output     = Chain(PrintMap(points = FALSE),
                     PerformanceMeasures))
```

```{r maxnet_plotting, eval = TRUE, echo = FALSE, dpi = 600, fig.show = "hide", fig.height = 6, fig.width = 9}

# plot the figure to disk, then plot in a separate chunk with a legend
par(mfrow = c(3, 2),
    mar = c(6, 5, 8, 5))

# loop through plotting them
for (i in 1:6) {
  fpath <- sprintf('figs/maxnet_comparison-%i.png', i)
  img <- brick(fpath)
  plotRGB(img)
  mtext(LETTERS[i],
        side = 3,
        line = -1.5,
        adj = 0)
}

```{r maxnet_plots, cache = FALSE, echo = FALSE, out.width = "450px", fig.cap = "Prediction of the distribution of the Carolina wren from 5 different models, produced by the workflow `maxnet\\_comparison'. A) MaxEnt without threshold features. B) MaxNet with default settings. C) MaxNet without regularisation. D) MaxNet with only linear features. E) MaxNet without regularisation and only linear features. F) Logistic regression. MaxNet with full features but no regularisation (C) gave the most local complexity, indicative of overfitting to the data. The models with only linear features (D-F) had a broader distribution, indicative of underfitting. Differences between MaxNet with linear features and no  regularisation (E) and logistic regression (F) are due to the simple downweighting applied to data in the former."}
knitr::include_graphics("figs/maxnet_plotting-1.png")
```

```{r maxent_aucs, eval = TRUE, echo = FALSE}
# AUCs to be embedded in the text below
perflist <- maxnet_comparison$report[seq(2, 12, by = 2)]
AUCs <- sapply(perflist, function(x) x$auc)
names(AUCs) <- c('MaxEnt',
                 'MaxNet',
                 'MaxNet (no regularization)',
                 'MaxNet (linear features)',
                 'MaxNet (linear features and no regularization)',
                 'Logistic regression')
AUCs <- round(AUCs, digits = 4)
```

The AUC statistics calculated against the presence-absence data were:
```{r results = 'asis', echo = FALSE}
cat(paste(names(AUCs), AUCs, collapse = '; '))
cat('.')
```

Readers interested in a more comprehensive comparison, or in exploring the response curves can easily reproduce and modify the analysis either by editing and re-running the above code block or by downloading the workflow object and using the `ChangeWorkflow` function.
